Stage-Splitting a Modal Language

### 2014-06-20 paper/semantics.tex
adding hats
iterating semantics explanation

A key attribute of the {\tt fexp} code example is the nesting of stage-\bbone\ and stage-\bbtwo\ expressions. Ordinary term evaluation eliminates outermost redexes first, 
however in the case that stage \bbone\ expressions are contained inside
stage \bbtwo\ ones (such as in the example above), this strategy conflicts with the precept of staged execution: that all stage \bbone\ code be evaluated before the evaluation of stage \bbtwo\ code. Therefore, a suitably staged dynamic semantics for \lang\ must evaluate
\emph{all} the stage \bbone\ subexpressions of a term before its stage
\bbtwo\ subexpressions.

To understand how this behavior can be achieved, consider evaluation of the stage-\bbone\ term below. (The term contains nested $\next$ and $\prev$ blocks.):
\begin{lstlisting} 
let x = (next {1+2}, 3+4) in
  next{ prev{#1 x} * prev{#1 x} * prev{hold(#2 x)} }
\end{lstlisting}
One option is to duplicate the contents of the first $\next$ expression, to produce
\begin{lstlisting}
(1+2) * (1+2) * 7
\end{lstlisting}

Here we are essentially treating $\next\{e\}$ as a stage-\bbone\ value, so long
as all the $\prev$ subterms of $e$ are reduced. This is the approach taken by
\cite{davies96}.

We take a different approach in our dynamic semantics, which is to let-bind the
contents of the first $\next$ expression to some variable (here, $x1$) and
duplicate that variable,
\begin{lstlisting} 
let x1 = 1+2 in
  x1 * x1 * 7
\end{lstlisting}

@@

To achieve this behavior, we must start by evaluating the binding for $x$.  We can't evaluate inside the
$\next$ statement, but must still substitute some result into the body of the
\texttt{let}. We replace the contents of $\next$ with a variable and make an
explicit substitution, which then floats up to the top of the containing
$\prev$:
\begin{lstlisting} 
[x1|->1+2]
let x = (next{x1}, 7) in
next{ prev{#1 x} * prev{#1 x} * prev{hold (#2 x)} }
\end{lstlisting}
We're now free to perform the stage-\bbone~substitution for $x$.  
\begin{lstlisting} 
[1+2/x1]
next{
	prev{#1 (next{x1}, 7)} *
	prev{#1 (next{x1}, 7)} *
	prev{hold (#2 (next{x1}, 7))}
}
\end{lstlisting}
To evaluate the remaining $\next$, we must first partially evaluate the body by finding all of the contained stage-\bbone~terms and reducing them:
\begin{lstlisting} 
[x1|->1+2]
	next{ x1 * x1 * 7 }
\end{lstlisting}
The final step in evaluating the outer $\prev$ is reifying the contained substitutions into let statements, yielding
\begin{lstlisting} 
let x1 = 1+2 in
  x1 * x1 * 7
\end{lstlisting}

@@

\TODO write the function here}
Given its stage \bbone{} argument $a:A$, we can evaluate the partially-applied
function:
$\cdot;\cdot\vdash f~a \mathop{\redonesym} [\xi,v]$.
The result is an environment $\xi$ and a partial value $v$ of type $\fut(B\to
C)$. Next, we reify this environment into a sequence of \verb|let|-bindings
enclosing $v$, via $\reify\xi{v}{f_a}$. A canonical forms theorem on $v$
ensures that the resulting $f_a$ has type $B\to C$ in \langTwo.
Finally, given a stage \bbtwo{} argument $b:B$, we can stage-\bbtwo{} evaluate
the ultimate result of the function, $f_a~b \mathop{\tworedsym} c$.

>>>>

A key attribute of the {\tt fexp} code example is the nesting of stage-\bbone\ and stage-\bbtwo\ expressions. Ordinary term evaluation eliminates outermost redexes first, 
however in the case that stage \bbone\ expressions are contained inside
stage \bbtwo\ ones (such as in the example above), this strategy conflicts with the precept of staged execution: that all stage \bbone\ code be evaluated before the evaluation of stage \bbtwo\ code. 
Accordingly, a suitably staged dynamic semantics for \lang\ must evaluate
\emph{all} the stage \bbone\ subexpressions of a term before its stage
\bbtwo\ subexpressions.

\subsection{Non-Duplicating First-Stage Evaluation}

To understand how this behavior can be achieved, consider evaluation of the stage-\bbtwo\ term below, which contains nested $\next$ and $\prev$ blocks:
\begin{lstlisting} 
prev{
  let x = (next {1+2}, 3+4) in
  next{ prev{#1 x} * prev{#1 x} * prev{hold(#2 x)} }
}
\end{lstlisting}
Our goal is to evaluate away all of the stage-\bbone\ portions of this code, but there is more than one reasonable way to do this.
One option, taken by \cite{davies96}, is to treat $\next\{\mathtt{1+2}\}$ as a value for the purposes of stage-\bbone~evaluation, 
consequently duplicating it during substitution for {\tt x}.  
This produces:
\begin{lstlisting}
(1+2) * (1+2) * 7
\end{lstlisting}

We take a different approach.  
Instead of duplicating the contents of the first $\next$ expression, we bind them to some variable (here, {\tt y}) and
duplicate only a reference to that variable.  This produces:
\begin{lstlisting} 
let y = 1+2 in
  y * y * 7
\end{lstlisting}

@@

Achieving this behavior mechanically requires us to resolve a contradiction:
we must substitute for \texttt{x} at stage \bbone, but we cannot evaluate inside the
$\next$ block bound to it. Our solution is to replace the $\next$ with a new variable and create an
explicit substitution (shown with a $\mapsto$) binding that variable to the $\next$'s contents.  
This substitution then floats up to the top of the containing $\prev$ block:
\begin{lstlisting} 
[y|->1+2]
let x = (yhat, 7) in
next{ prev{#1 x} * prev{#1 x} * prev{hold (#2 x)} }
\end{lstlisting}
Note that the variable left behind is rendered with a hat.  By treating this {\em hatted variable} as a value, we're now free to perform the stage-\bbone~substitution for {\tt x}.  
\begin{lstlisting} 
[x1|->1+2]
next{
	prev{#1 (yhat, 7)} *
	prev{#1 (yhat, 7)} *
	prev{hold (#2 (next{x1}, 7))}
}
\end{lstlisting}
To evaluate the remaining $\next$, we must first partially evaluate the body by finding all of the contained stage-\bbone~terms and reducing them:
\begin{lstlisting} 
[x1|->1+2]
	next{ x1 * x1 * 7 }
\end{lstlisting}
The final step in evaluating the outer $\prev$ is reifying the contained substitutions into let statements, yielding
\begin{lstlisting} 
let x1 = 1+2 in
  x1 * x1 * 7
\end{lstlisting}

@@

%\begin{lstlisting} 
%let adjust (f : $int * int -> $int) =
%	fn (p : int) => 
%		next{
%			fn (b : int) => 
%				prev{f (next {b}, p)}
%		}	
%\end{lstlisting}
\TODO code written, but listings don't work in footnotes! commented out in source
}
Given its stage \bbone{} argument $a:A$, we can evaluate the partially-applied
function:
$\cdot;\cdot\vdash f~a \mathop{\redonesym} [\xi,v]$.
The result is an environment $\xi$ and a partial value $v$ of type $\fut(B\to
C)$. Next, we reify this environment into a sequence of \verb|let|-bindings
enclosing $v$, via $\reify\xi{v}{f_a}$. A canonical forms theorem on $v$
ensures that the resulting $f_a$ has type $B\to C$ in \langTwo.
Finally, given a stage \bbtwo{} argument $b:B$, we can stage-\bbtwo{} evaluate
the ultimate result of the function, $f_a~b \mathop{\tworedsym} c$.
